Relationships in the extremes and their influence on competition and coexistence

### 2021-09-14 paper/Paper.Rnw
continuing edits to Intro

%Intro to competition, coexistence, MCT, etc.
Competition and the maintenance of biodiversity are research topics at the core of modern ecology 
\citep{MacArthur_1958, Hutchinson_1961, Schmida_1984}.
The competitive exclusion principle states 
that two species competing for the same limiting resource cannot coexist at constant densities - one will instead outcompete 
the other and drive it to extinction \citep{Hardin_1960}.
Modern coexistence theory [MCT; \cite{Chesson_1994,Chesson_2000,Barabas_2018}] is a 
framework for understanding how species nevertheless coexist, and biodiversity in maintained, in real ecosystems. 
Coexistence occurs through stabilizing factors and niche 
differentiation, in spite of the conclusions drawn for idealized scenarios by the competitive exclusion principle. 
MCT breaks down coexistence into the contributions of various mechanisms, with names such as `relative nonlinearity
in competition' and `storage effects' \citep{Chesson_1994} which will be familiar to many readers. The MCT has been
applied to multiple systems, including annual plants \citep{Angert_2009, Hallett_2019}, perennial plants \citep{Adler_2010, Adler_2006}, 
tropical trees \citep{Usinowicz_2012}, plankton \citep{Narawani_2013, Caceres_1997}, and corals \citep{Alvarez-Noriega_2020}. 
Although the MTC is mathematically complex to most ecologists, recent extensions make it more accessible \citep{Ellner_2016,Ellner_2019}.

%Intro to ATAs
Every ecological discipline involves the study of relationships among biological and environmental variables,
but standard methods for examining the association between two variables provide only a small 
portion of the overall information content of the
relationship \citep{Ghosh_2020_AER,Ghosh_2020_Eco,Ghosh_2020_ecolevol,Ghosh_2021}. For instance, 
the three pairs of variables in figure \ref{fig:noise} all have the same
Pearson correlation, but nevertheless demonstrate very different types of association. Other metrics 
of association are also used, in addition to correlations
(e.g., various common forms of regression), but correlations are extremely common, and other common
methods also reveal only a small portion of the information content of the relationship between variables \citep{Nelsen2006}.
The variables in figure \ref{fig:noise}A,C 
exhibit what we have previously referred to as ``asymmetric tail association'' [ATA; \cite{Ghosh_2020_AER}], i.e., 
they are more strongly related to each
other in the left (figure \ref{fig:noise}A, termed ``left-tail association'') or right (figure \ref{fig:noise}C,
termed ``right-tail association'') portions of the variables' distributions. Although ATAs and related ideas had previously been very seldom 
considered in ecology and related areas (but see \citep{Anderson_2019, deValpine_2014, Popovic_2019}), 
our recent work has shown that, in fact, ATAs are extremely common in ecology \citep{Ghosh_2020_AER}, and can influence 
many ecological phenomena, including extinction risk \citep{Ghosh_2020_Eco}, Taylor's law \citep{Ghosh_2020_AER}, 
synchrony, and the temporal stability of ecosystem functioning \citep{Ghosh_2020_ecolevol,Ghosh_2021}.

@@

what storage effects are \citep{Chesson_1994,Chesson_2012,Ellner_2016,Barabas_2018} and we explain why ATAs may be relevant to them.

@@

whichever species is currently rare; this is the characteristic that separates so-called \emph{stabilizing} mechanisms of coexistence from
\emph{equalizing} mechanisms, which instead reduce average fitness differences between comeptitors \citep{Chesson_2000}.
Storage effects can occur if three conditions are met \citep{Chesson_1994}. First, there must be positive covariation between the quality of the environment (E)

@@

in times when the environment is poor; environmentally insensitive adult stages and dormant stages such as seed banks are
example mechanisms \citep{Chesson_2012}. 
When present, these factors combine so that EC covariance provides a relative benefit to whatever species is currently rare.
Storage effects are quantified by measuring to what extent EC covariance contributes to differences between 
growth rates of species when they are at low ``invader'' densities, and growth rates of the abundant ``resident'' species \citep{Chesson_1994,Ellner_2016}. 
Thus ATAs may influence storage effects because storage effects involve EC covariation. ATAs between E and C may,
in principle, mediate the strength of storage effects, though to our knowledge this possibility has not previously been investigated.
%***

@@

%Conversely, EC covariance is lesser for the low density species, leading to large fluctuations in population growth rate in both directions. 

@@

@@

Environmental fluctuations are directly impacted by climate change through changes in means and increased variance \citep{Hansen_2012}, influenced by extreme climatic events as they increase in frequency, intensity, duration, and exent \citep{Meehl_2004, Ummenhofer_2017}. 
The rampage of extreme events such as heat-waves, floods, hurricanes, fires, etc. are a product of climate change and have implications for coexistence and competition. Changes in means can be attributed to extreme events; however changes in means are moreso a result of gradual increases or decreases over time in an environmental variable. 
Extreme events push environmental variables to their extremes, increasing variance. Effects on competition and coexistence due to climate change are more widley researched through studing changes in the mean of environmental variables. 
For example, decreased growth season precipitation can decrease the competitive ability of dominant forest tree species \citep{Clark_2011}. This may be detrimental if the dominant species is crucial to the ecosystem or it may not be as rare species can increase in diversity while dominant species become excluded \citep{Mariotte_2013}. Conversley, changes in environmental means can increase fitness of a dominant species and in turn reduce community diversity, typical of invasive species \citep{Vitousek_1997}. \cite{Descombes_2020}, through experimental manipulation, found that range shifts, as a product of climate change, favored coexistence between alpine plants while \cite{Usinowicz_2021} predict increases in competitive pressures as the climate change induced range shifts lead populations to interact with new conspecifics. Similarly, \cite{diPaola_2012} predict the collapse of coexistence between Mediterranean tree species as rainfall levels are forecasted to decrease.    
In contrast, increased environmental variance has been less commmonly studied; however, more research on the effect of increased variance has been urged and received recent focus through the study of extreme climatic events \citep{Jentsch_2007, Smith_2011}. Extreme drought and heat waves impact competitive output and community composition of grassland plants\citep{White_2000, White_2001, Hoover_2014}. However, complete exclusion of dominant species due to extreme events is uncommon \citep{Lloret_2012}. Morever, in regard to invasive species, increased climate and resource variability allowed for greater success of invasive plants in terms of density and competitive equilibrium \citep{Parepa_2013}. Extreme events constitute large disturbances, quantitatively amounting to large fluctuations in environmental drivers and their ecological responses \citep{Smith_2011}. Although, increased variability of environmental fluctuations is a newer consideration for studying environmental changes; environmental variability has been analytically shown to be important for classical coexistence \citep{Chesson_1981, Warner_1985}. Moreover, increased variability has been found to promote coexistence in plants communities \citep{Adler_2006,  Angert_2009}, but disrupt coexistence in birds \citep{Stenseth_2002, Stenseth_2015}. This present study has applications to extreme events and climate change as we consider relationships in their extremes, potentially caused by extreme events, and their implications for competition and coexistence.

Extreme events can have important ecological impacts for controphic competing communities where species growth and the maintenance of biodiversity depends on distinct temporal events/dynamics. If extreme events are more temporally extensive than moderate disturbances, then it is more likely that two species will be influenced by the same events. Both species may become more alike (correlated), an important consideration for competition and coexistence. For example, an extreme flood event that lasts across the growth season of two different species and increases recruitment in both species will cause recruitment of the two species to become more correlated in the upper tail end of the distribution. A subsequent change in competitive ability and coexistence is not just attributed to the extreme event alone, but from the resulting extreme relationship between species (ATA). Though previous literature considers impacts of extreme events, change in means, and increased variability on species coexistence and competition and community dynamics in gerneral (see above), they do not directly consider ATA between species response caused by environmental changes -- extreme relationships caused by extreme events (but see \cite{Ghosh_2021} for impact of ATA on community aggregate properties).

@@

%move this paragraph to discussion
%Previously, storage effects were described as the contribution of covariance between the environment and competition to coexistence \citep{Chesson_2012, Ellner_2016, Ellner_2019}. 
%However, the description of storage effects as the covariance between drivers is a (potentially dangerous) misnomer in terminology, and we argue that storage effects, instead, should be described as the contribution of the relationship between drivers. 
%All aspects and contributing values of the relationship between environmental response and competitive pressure are not fully captured by only the covariance between them because a measure of covariance does not capture the strucutre of associativity throuhgout the distribution of the relationship and any asymmetries there may be. 
%Thus, aligning with the decomposition presented in \cite{Ellner_2019}, we propose storage effect can be broken down further into the contribution of ATA [to storage effect/coexistence] and the contribution of symmetric association [to storage effect/coexistence].
We have demonstrated that ATAs exist and are common in nature \citep{Ghosh_2020_AER}, and we have elucidated how ATAs between environmental response variables might arise via extreme events. ATA in environmental response directly feeds into an ATA effect on the EC covariance contribution to the storage effect mechanism of coexistence. Thus we seek to explore the consequences of ATAs on coexistence, competition, and the maintenance of biodiversity.
This paper has the following goals. 
G1) To augment the modern theory of coexistence (MTC) to show how ATA plays a role. 
The MTC decomposes a measure of coexistence into components due to each of several mechanisms. 
We address how ATA contributes to these components. 
We here examine mechanisms of coexistence that apply in models exhibiting solely temporal variation; spatial variation can be considered in future work.
G2) To explore, using a simple model, the relative magnitudes of the contributions to coexistence that come
from ATAs compared to other mechanisms. This is to help illuminate whether ATA contributions to coexistence 
are likely to be negligible or meaningfully large. 
%G3) To identify and define the ATA rescue or ATA exlusion set of trait space range for which species will coexist or become competitively excluded in our model. 
%***DAN: This is a good additional goal. We will either have to have already defined the "ATA rescue and ATA exlusion sets of trait space"
%or else readers won't understand this as it was written, above (now commented). These are complex concepts, so it may be better to 
%save formal definitions for later and reword this. Here's a try at rewording: 
G3) To explore, using our simple model, whether species can coexist with
ATAs but not without, or vice versa. This is to begin to illuminate whether ATA contributions to biodiversity maintenance 
may be negligible or meaninfully large.
G4) To provide an empirical example of species coexistence, and to highlight the role ATAs play in coexistence for the example.
In the Discussion, we revisit the potential that climatic changes will influence
the roles of ATAs in species coexistence and the maintenance of biodiversity.
Overall, our study presents a new mechanism of coexistence and biodiversity maintenance which may be influenced by climate change, 
and a means for understanding its importance for real systems. This is not a good ending sentence.

@@

>>>>

%Intro to competition, coexistence, MCT, etc.
Competition and the maintenance of biodiversity are research topics at the core of modern ecology 
\citep{MacArthur_1958, Hutchinson_1961, Schmida_1984}.
The competitive exclusion principle states 
that two species competing for the same limiting resource cannot coexist at constant densities - one will instead outcompete 
the other and drive it to extinction \citep{Hardin_1960}.
Modern coexistence theory [MCT; \cite{Chesson_1994,Chesson_2000,Barabas_2018}] is a 
framework for understanding how species nevertheless coexist, and biodiversity in maintained, in real ecosystems. 
Coexistence occurs through stabilizing factors and niche 
differentiation, in spite of the conclusions drawn for idealized scenarios by the competitive exclusion principle. 
MCT breaks down coexistence into the contributions of various mechanisms, with names such as `relative nonlinearity
in competition' and `storage effects' \citep{Chesson_1994} which will be familiar to many readers. The MCT has been
applied to multiple systems, including annual plants \citep{Angert_2009, Hallett_2019}, perennial plants \citep{Adler_2010, Adler_2006}, 
tropical trees \citep{Usinowicz_2012}, plankton \citep{Narawani_2013, Caceres_1997}, and corals \citep{Alvarez-Noriega_2020}. 
Although the MTC is mathematically complex to many ecologists, recent extensions make it more accessible \citep{Ellner_2016,Ellner_2019}.

%Intro to ATAs
Every ecological discipline involves the study of relationships among biological and environmental variables,
but standard methods for examining the association between two variables provide only a small 
portion of the overall information content of the
relationship \citep{Ghosh_2020_AER,Ghosh_2020_Eco,Ghosh_2020_ecolevol,Ghosh_2021}. For instance, 
the three pairs of variables in figure \ref{fig:noise} all have the same
Pearson correlation, but nevertheless demonstrate very different types of association. Other metrics 
of association are also used, in addition to correlations
(e.g., various common forms of regression), but correlations are extremely common, and other common
methods also reveal only a small portion of the information content of the relationship between variables \citep{Nelsen2006}.
The variables in figure \ref{fig:noise}A,C 
exhibit what we have previously referred to as ``asymmetric tail association'' [ATA; \cite{Ghosh_2020_AER}], i.e., 
they are more strongly related to each
other in the left (figure \ref{fig:noise}A, termed ``left-tail association'') or right (figure \ref{fig:noise}C,
termed ``right-tail association'') portions of the variables' distributions. Although ATAs 
and related ideas had previously been very seldom 
considered in ecology and related areas (but see \citealp{Anderson_2019, deValpine_2014, Popovic_2019}), 
our recent work has shown that, in fact, ATAs are extremely common in ecology \citep{Ghosh_2020_AER}, and can influence 
many ecological phenomena, including extinction risk \citep{Ghosh_2020_Eco}, Taylor's law \citep{Ghosh_2020_AER}, 
synchrony, and the temporal stability of ecosystem functioning \citep{Ghosh_2020_ecolevol,Ghosh_2021}.

@@

what storage effects are \citep{Chesson_1994,Chesson_2012,Ellner_2016,Barabas_2018}, and we explain why ATAs may be relevant to them.

@@

whichever species is currently rare; this is the characteristic that make storage effects a so-called \emph{stabilizing} mechanism 
of coexistence \citep{Chesson_2000}.
Storage effects can occur if three conditions are met \citep{Chesson_1994}. First, there must be positive 
covariation between the fluctuating quality of the environment (E)

@@

in times when the environment is poor; dormant stages such as are in seed banks are an
example buffering mechanisms \citep{Chesson_2012}. 
When present, these three factors combine so that EC covariance provides a relative benefit to whatever species is currently rare.
Importantly, storage effects are, in fact, quantified by measuring to what extent EC covariance contributes to differences between 
invader and resident growth rates \citep{Chesson_1994,Ellner_2016}.
Thus ATAs may, in principle, influence storage effects simply because storage effects hinge on EC covariation, so ATAs 
between between E and C may mediate the strength of storage effects.
%***Old text below, kept for now, but should be pretty much replaced by the above

@@

%Conversely, EC covariance is lesser for the low density species, leading to large fluctuations in population 
%growth rate in both directions. 

@@

%***DAN: I think the above paragraphs are pretty solid. I am still working on the below.

@@

%***DAN: Jasmin, there are a lot of disjointed sentences in the below paragraph, at least they appear disjointed to me.
%Can you please say in a comment right below this one what the topic sentence is? As usual, if you are having trouble 
%determining what the topic sentence it, it may mean you don't have one, and you need to determine what you're doing 
%with this paragraph and make a topic sentence. 
Environmental fluctuations are directly impacted by climate change through changes in means and increased variance \citep{Hansen_2012}, influenced by extreme climatic events as they increase in frequency, intensity, duration, and exent \citep{Meehl_2004, Ummenhofer_2017}. 
The rampage of extreme events such as heat-waves, floods, hurricanes, fires, etc. are a product of climate change and have implications for coexistence and competition. Changes in means can be attributed to extreme events; however changes in means are moreso a result of gradual increases or decreases over time in an environmental variable. 
Extreme events push environmental variables to their extremes, increasing variance. Effects on competition and coexistence due to climate change are more widley researched through studing changes in the mean of environmental variables. 
For example, decreased growth season precipitation can decrease the competitive ability of dominant forest tree species \citep{Clark_2011}. This may be detrimental if the dominant species is crucial to the ecosystem or it may not be as rare species can increase in diversity while dominant species become excluded \citep{Mariotte_2013}. Conversley, changes in environmental means can increase fitness of a dominant species and in turn reduce community diversity, typical of invasive species \citep{Vitousek_1997}. \cite{Descombes_2020}, through experimental manipulation, found that range shifts, as a product of climate change, favored coexistence between alpine plants while \cite{Usinowicz_2021} predict increases in competitive pressures as the climate change induced range shifts lead populations to interact with new conspecifics. Similarly, \cite{diPaola_2012} predict the collapse of coexistence between Mediterranean tree species as rainfall levels are forecasted to decrease.    
In contrast, increased environmental variance has been less commmonly studied; however, more research on the effect of increased variance has been urged and received recent focus through the study of extreme climatic events \citep{Jentsch_2007, Smith_2011}. Extreme drought and heat waves impact competitive output and community composition of grassland plants\citep{White_2000, White_2001, Hoover_2014}. However, complete exclusion of dominant species due to extreme events is uncommon \citep{Lloret_2012}. Morever, in regard to invasive species, increased climate and resource variability allowed for greater success of invasive plants in terms of density and competitive equilibrium \citep{Parepa_2013}. Extreme events constitute large disturbances, quantitatively amounting to large fluctuations in environmental drivers and their ecological responses \citep{Smith_2011}. Although, increased variability of environmental fluctuations is a newer consideration for studying environmental changes; environmental variability has been analytically shown to be important for classical coexistence \citep{Chesson_1981, Warner_1985}. Moreover, increased variability has been found to promote coexistence in plants communities \citep{Adler_2006,  Angert_2009}, but disrupt coexistence in birds \citep{Stenseth_2002, Stenseth_2015}. This present study has applications to extreme events and climate change as we consider relationships in their extremes, potentially caused by extreme events, and their implications for competition and coexistence.

%***DAN: I am also having a hard time identifying the topic sentence of this paragraph, can you tell me which one it is, please,
%in a comment right under this one?
Extreme events can have important ecological impacts for controphic 
%***DAN: I don't know what "controphic" means and when I google it I get nothing. Perhaps you intended a different word?
competing communities where species growth and the maintenance of biodiversity depends on distinct temporal events/dynamics. If extreme events are more temporally extensive than moderate disturbances, then it is more likely that two species will be influenced by the same events. Both species may become more alike (correlated), an important consideration for competition and coexistence. For example, an extreme flood event that lasts across the growth season of two different species and increases recruitment in both species will cause recruitment of the two species to become more correlated in the upper tail end of the distribution. A subsequent change in competitive ability and coexistence is not just attributed to the extreme event alone, but from the resulting extreme relationship between species (ATA). Though previous literature considers impacts of extreme events, change in means, and increased variability on species coexistence and competition and community dynamics in gerneral (see above), they do not directly consider ATA between species response caused by environmental changes -- extreme relationships caused by extreme events (but see \cite{Ghosh_2021} for impact of ATA on community aggregate properties).

@@

We have demonstrated that ATAs exist and are common in nature \citep{Ghosh_2020_AER}, and we have elucidated how ATAs between environmental response variables might arise via extreme events. ATA in environmental response directly feeds into an ATA effect on the EC covariance contribution to the storage effect mechanism of coexistence. Thus we seek to explore the consequences of ATAs on coexistence, competition, and the maintenance of biodiversity.
This paper has the following goals. 
G1) To augment the modern theory of coexistence (MTC) to show how ATA plays a role. 
The MTC decomposes a measure of coexistence into components due to each of several mechanisms. 
We address how ATA contributes to these components. 
We here examine mechanisms of coexistence that apply in models exhibiting solely temporal variation; spatial variation can be considered in future work.
G2) To explore, using a simple model, the relative magnitudes of the contributions to coexistence that come
from ATAs compared to other mechanisms. This is to help illuminate whether ATA contributions to coexistence 
are likely to be negligible or meaningfully large. 
%G3) To identify and define the ATA rescue or ATA exlusion set of trait space range for which species will coexist or become competitively excluded in our model. 
%***DAN: This is a good additional goal. We will either have to have already defined the "ATA rescue and ATA exlusion sets of trait space"
%or else readers won't understand this as it was written, above (now commented). These are complex concepts, so it may be better to 
%save formal definitions for later and reword this. Here's a try at rewording: 
G3) To explore, using our simple model, whether species can coexist with
ATAs but not without, or vice versa. This is to begin to illuminate whether ATA contributions to biodiversity maintenance 
may be negligible or meaninfully large.
G4) To provide an empirical example of species coexistence, and to highlight the role ATAs play in coexistence for the example.
In the Discussion, we revisit the potential that climatic changes will influence
the roles of ATAs in species coexistence and the maintenance of biodiversity.
Overall, our study presents a new mechanism of coexistence and biodiversity maintenance which may be influenced by climate change, 
and a means for understanding its importance for real systems. This is not a good ending sentence.

@@

%Need something like the following, maybe
%Previously, storage effects were described as the contribution of covariance between the environment and competition to coexistence \citep{Chesson_2012, Ellner_2016, Ellner_2019}. 
%However, the description of storage effects as the covariance between drivers is a (potentially dangerous) misnomer in terminology, and we argue that storage effects, instead, should be described as the contribution of the relationship between drivers. 
%All aspects and contributing values of the relationship between environmental response and competitive pressure are not fully captured by only the covariance between them because a measure of covariance does not capture the strucutre of associativity throuhgout the distribution of the relationship and any asymmetries there may be. 
%Thus, aligning with the decomposition presented in \cite{Ellner_2019}, we propose storage effect can be broken down further into the contribution of ATA [to storage effect/coexistence] and the contribution of symmetric association [to storage effect/coexistence].
